---
title: "A1 Summary with 2019 Method"
author: "Samuel Ingloff-Richards"
date: "2025-09-25"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r}
library(ggplot2)
library(tidyr)
```

# Preview data and matrices

```{r}
A1 <- read.csv("output/A1/A1_adjusted.csv")
nee22 <- read.csv("data/Global predictors of language endangerment and the future of linguistic diversity Data 2.csv")  # (3)
load("output/A1/phylomatrix.RData")
load("output/A1/spmatrix.RData")
load("output/A1/ml_model.RData")
```

## A1 Dataset

```{r}
dim(A1)
print(A1[1:6,],width = 150)
```

## Overview

-   769 languages in dataset

## Spatial Distance (3) Matrix

```{r}
paste(dim(spmatrix))
paste("Is symmetric? ", isSymmetric(spmatrix))
print(spmatrix[1:6,1:6])
```

## Phylogenetic (3) Matrix

```{r}
print(dim(phylomatrix))
paste("Is symmetric? ", isSymmetric(phylomatrix))
print(phylomatrix[1:6,1:6])
```

# Variable Table

| Variable | Definition | Example (abu, abur1243, Abure) | Source |
|----|----|----|----|
| Phoneme.Inventory.Size | Number of phonemes in a language inventory. | 34 | (4) |
| L1_pop | Number of L1 speakers for a language | 55100 | (3) |

# Method Overview

1.  Merge data from (3) and (4).

2.  Remove NA entries in merged data.

3.  Remove duplicate languages (due to Phoible having multiple inventories for same language, or many Glottocodes for ISO639-3).

4.  Adjust size of matrices with merged data to only shared languages, ending up with 769 languages.

5.  Log transform variables

6.  Find optimal p vector (encoding weighting of spatial and phylogenetic matrices) using 2019 (2) method, on only model: `Phoneme.Inventory.Size ~ L1_pop`

7.  Using 2019 (2) method and optimal p vector from step 6, find generalised least squares (GLS) maximum likelihood solution for each model.

# Results

## Phoneme.Inventory.Size \~ L1_pop scatterplot grouped by region.

```{r}
# Plot data
phoneme_L1_scatter <- ggplot(A1, aes(x = log(L1_pop+0.5), y = log(Phoneme.Inventory.Size), color = factor(region, levels = unique(region)))) + 
  geom_point() + 
  # stat_ellipse(geom = "polygon", alpha = 0.2, aes(fill = group)) + 
  labs(x = "Log(L1_pop+0.5)", y = "Log(Phoneme.Inventory.Size)", color = "Region") +
  theme_minimal()

# Save graph
ggsave("output/A1/phoneme_to_L1_scatter.png", plot = phoneme_L1_scatter, width = 15, height = 15, dpi = 300)

# Show graph
print(phoneme_L1_scatter)
```

## Testing for region sampling bias

```{r}
# number of repetitions
n_rep = 1e4

# Get region names
region_names <- unique(A1$region)
region_names <- sort(region_names)

# Find number of languages in each region for A1
A1_region_tally <- as.data.frame(table(A1$region))
A1_region_tally <- A1_region_tally[match(region_names, A1_region_tally$Var1), ]

region_sample_tallies <- data.frame(matrix(ncol = length(region_names), nrow = n_rep))
colnames(region_sample_tallies) <- region_names # Set column names

# Repeated n_rep times, randomly sample without replacement 
for(i in 1:n_rep){
  
  # Get region sample
  region_sample <- sample(nee22$region, size = length(A1$region), replace = FALSE)
  
  # Current region tally.
  region_sample_tally = table(region_sample)
  
  # Add current tally to all tally data
  region_sample_tallies[i, names(region_sample_tally)] <- as.numeric(region_sample_tally)
   
}

# Reshape the dataframe to long format
region_sample_tallies_long <- pivot_longer(region_sample_tallies, everything(), names_to = c("Group"), values_to = 'Value')

# Set 'Group' as a factor with levels in the desired order, this is to ensure that ggplot doesn't alphabetically sort the regions.
region_sample_tallies_long$Group <- factor(region_sample_tallies_long$Group, levels = region_names)

region_box <- ggplot(region_sample_tallies_long, aes(x = Group, y = Value)) +
  geom_boxplot(varwidth = TRUE) +
  geom_point(data = A1_region_tally, 
             aes(x = region_names, y = Freq), 
             color = "red",
             size = 2) + 
  ylab("Number of Languages Sampled") +
  xlab("Region") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 10),
        axis.title.x.bottom = element_text(size = 10),
        axis.text.y = element_text(size = 10),
        axis.title.y.left = element_text(size = 10))

# Save graph
ggsave("output/A1/region_sampling.png", plot = region_box, width = 15, height = 15, dpi = 300)

print(region_box)

```

*Randomly sampled languages from (3) data for their region values, and compared sampling distribution with observed region values in A1 data.*

# Phoneme.Inventory.Size across regions

```{r}
box_phoneme_regions <- ggplot(A1, aes(
  x = factor(region),
  y = Phoneme.Inventory.Size
)) +
  geom_boxplot(position = position_dodge(width = 0.8)) +
  labs(
    x = "region",
    y = "Phoneme.Inventory.Size"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(box_phoneme_regions)

# Save graph
ggsave("output/A1/phoneme_region_box.png", plot = box_phoneme_regions, width = 15, height = 15, dpi = 300)
```

# 

## Results using 2019 (2) method

```{r}
print("Coefficients and p-values")
summary(ml_model)$tTable
paste("Log likelihood: ", summary(ml_model)$logLik)
paste("BIC: ", summary(ml_model)$BIC)
```

# References

1.  Bromham, L., Yaxley, K.J. & Cardillo, M. Islands are engines of language diversity. *Nat Ecol Evol* **8**, 1991–2002 (2024). <https://doi.org/10.1038/s41559-024-02488-4>
2.  Hua, X., Greenhill, S.J., Cardillo, M. *et al.* The ecological drivers of variation in global language diversity. *Nat Commun* **10**, 2047 (2019). <https://doi.org/10.1038/s41467-019-09842-2>
3.  Bromham, L., Dinnage, R., Skirgård, H. *et al.* Global predictors of language endangerment and the future of linguistic diversity. *Nat Ecol Evol* **6**, 163–173 (2022). <https://doi.org/10.1038/s41559-021-01604-y>
4.  Cormac Anderson, Tiago Tresoldi, Simon J Greenhill, Robert Forkel, Russell Gray, Johann-Mattis List, Variation in phoneme inventories: quantifying the problem and improving comparability, *Journal of Language Evolution*, Volume 8, Issue 2, July 2023, Pages 149–168, <https://doi.org/10.1093/jole/lzad011>
