---
title: "Anderson Summary with 2019 Method"
author: "Samuel Ingloff-Richards"
date: "2025-09-24"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Preview data and matrices

```{r}
anderson <- read.csv("output/anderson/anderson_adjusted.csv")
load("output/anderson/phylomatrix.RData")
load("output/anderson/spmatrix.RData")
ml_data <- read.csv("output/anderson/ml_data.csv")
```

## Anderson (4) Dataset

```{r}
dim(anderson)
print(anderson[1:6,],width = 150)
```

## Spatial Distance (3) Matrix

```{r}
paste(dim(spmatrix))
paste("Is symmetric? ", isSymmetric(spmatrix))
print(spmatrix[1:6,1:6])
```

## Phylogenetic (3) Matrix

```{r}
print(dim(phylomatrix))
paste("Is symmetric? ", isSymmetric(phylomatrix))
print(phylomatrix[1:6,1:6])
```

# Variable Table

| Variable | Definition | Example (abu, abur1243, Abure) | Source |
|----|----|----|----|
| Phoneme.Inventory.Size | Number of phonemes in a language inventory. | 34 | (4) |
| L1_pop | Number of L1 speakers for a language | 55100 | (3) |
| bordering_language_richness | Number of languages bordering a given language | 7 | (3) |
| altitude_range | Altitude at which language is located | 180.58469 | (3) |
| Island.Endemic | If island is endemic to an island (see definition for in (2) | 0 (False) | (1) |
| Distance.to.Continent | Distance (in km) language is from nearest continent | 0 | (1) |
| Distance.to.Mainland | Distance (in km) language is from nearest mainland | 0 | (1) |
| Range.Size..km2. | Geographic range (in km\^2) that language spans. | 1116.49713 | (1) |

# Method Overview

1.  Merge data from (1), (3) and (4).

2.  Remove NA entries in merged data.

3.  Remove duplicate languages (due to Phoible having multiple inventories for same language, or many Glottocodes for ISO639-3).

4.  Adjust size of matrices with merged data to only shared languages, ending up with 762 languages.

5.  Log transform variables. Although zero values in `Distance.to.Continent` and `Distance.to.Mainland` were not log transformed, e.g:

`anderson$Distance.to.Mainland[anderson$Distance.to.Mainland != 0] <- log( anderson$Distance.to.Mainland[anderson$Distance.to.Mainland != 0])`

6.  Find optimal p vector (encoding weighting of spatial and phylogenetic matrices) using 2019 (2) method, on interceot only model: `Phoneme.Inventory.Size ~ 1`
7.  Create all model combinations for given predictors, here 128 models (including intercept only model).
8.  Using 2019 (2) method and optimal p vector from step 6, find generalised least squares (GLS) maximum likelihood solution for each model.
9.  Sort models based on BIC values.

# Results using 2019 (2) Method

Show top 10 models

```{r}
print(ml_data[1:10,],width = 150)
```

*128 models ranked by BIC. logLik = Log likelihood, MSE = Mean Square Error (of model residuals). NA indicates that predictor was not used in the model, all NA model is the intercept model: Phoneme.Inventory.Size \~ 1. Predictor columns show their respective coefficient values, if "\*\*\*" is appended to a coefficient it indicates that the predictor in this model has a p-value below 0.05.*

## Summary

`L1_pop` is the only predictor across all models which has p-values below 0.05. But the predictor only seems to be weakly related to `Phoneme.Inventory.Size`.

# References

1.  Bromham, L., Yaxley, K.J. & Cardillo, M. Islands are engines of language diversity. *Nat Ecol Evol* **8**, 1991–2002 (2024). <https://doi.org/10.1038/s41559-024-02488-4>
2.  Hua, X., Greenhill, S.J., Cardillo, M. *et al.* The ecological drivers of variation in global language diversity. *Nat Commun* **10**, 2047 (2019). <https://doi.org/10.1038/s41467-019-09842-2>
3.  Bromham, L., Dinnage, R., Skirgård, H. *et al.* Global predictors of language endangerment and the future of linguistic diversity. *Nat Ecol Evol* **6**, 163–173 (2022). <https://doi.org/10.1038/s41559-021-01604-y>
4.  Cormac Anderson, Tiago Tresoldi, Simon J Greenhill, Robert Forkel, Russell Gray, Johann-Mattis List, Variation in phoneme inventories: quantifying the problem and improving comparability, *Journal of Language Evolution*, Volume 8, Issue 2, July 2023, Pages 149–168, <https://doi.org/10.1093/jole/lzad011>
