---
title: "NEE24_Analysis"
output:
  pdf_document: default
  html_document: default
date: "2025-09-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyr)
library(nloptr)
library(nlme)
library(parallel)
```

# Data

## Loading Phylogenetic and Spatial Distance Matrices

```{r}
phylomatrix <- read.csv("data/distance_matrices/phylogenetic_covariance_matrix.csv",
                        check.names = FALSE)
rownames(phylomatrix) <- phylomatrix[,1]
phylomatrix <- phylomatrix[,-1]
phylomatrix <- as.matrix(phylomatrix)
spmatrix <- read.csv("data/distance_matrices/spatial_distance_matrix.csv",
                     check.names = FALSE)
rownames(spmatrix) <- spmatrix[,1]
spmatrix <- spmatrix[,-1]
spmatrix <- as.matrix(spmatrix)
```

### Preview Matrices

```{r}
print(phylomatrix[1:5,1:5])
print(spmatrix[1:5,1:5])
```

## Loading NEE24 (1)

```{r}
nee24 <- read.csv("data/Islands are engines of language diversity data.csv",
                  stringsAsFactors = FALSE)
```

### NEE24 Cleanup

```{r}
# --- Convert variable types ---
nee24$Island.Endemic <- as.numeric(nee24$Island.Endemic)

# --- Log transform variables ---
nee24$Phoneme.Inventory.Size <- log(nee24$Phoneme.Inventory.Size) 
nee24$Range.Size..km2. <- log(nee24$Range.Size..km2.)
nee24$L1.Population <- log(nee24$L1.Population + 0.5)
nee24$Distance.to.Mainland <- log(nee24$Distance.to.Mainland + 0.5)
nee24$Distance.to.Continent <- log(nee24$Distance.to.Continent + 0.5)

```

### Preview data

```{r}
head(nee24)
summary(nee24)
```

## Loading NEE22 (3)

```{r}
nee22 <- read.csv("data/Global predictors of language endangerment and the future of linguistic diversity Data 2.csv",
                  stringsAsFactors = FALSE)
```

### NEE22 Cleanup

```{r}
# Convert Documention variable from string to integer
nee22$documentation[nee22$documentation == "little or none"] <- "0"
nee22$documentation[nee22$documentation == "basic"] <- "1"
nee22$documentation[nee22$documentation == "detailed"] <- "2"
nee22$documentation <- as.integer(nee22$documentation)

# --- Log transform variables ---
nee22$L1_pop <- log(nee22$L1_pop + 0.5)
nee22$bordering_language_richness <- log(nee22$bordering_language_richness + 0.5)
nee22$altitude_range <- log(nee22$altitude_range + 0.5)
```

### Preview data

```{r}
head(nee22)
summary(nee22)
```

## Loading phoible (4) 

```{r}
var_phon_inv_spec_data <- read.csv("Data/cldf-datasets-inventory-study/PHOIBLE-data.csv")
var_phon_inv_glot_data <- read.csv("Data/cldf-datasets-inventory-study/phoible/cldf/languages.csv")

phoible <- merge(
  var_phon_inv_spec_data[, c("Glottocode", "Sounds", "Latitude", "Longitude")],
  var_phon_inv_glot_data[, c("ISO639P3code", "Glottocode")], by.x = "Glottocode", by.y = "Glottocode", all.x = TRUE)
```

### Phoible Cleanup

```{r}
# --- Remove duplicate glottocode entries
print("Duplicate Glottocode entries:")
glotto_duplicates <- duplicated(phoible$Glottocode)
print(phoible$Glottocode[glotto_duplicates])
phoible <- phoible[!unlist(glotto_duplicates),]
paste("Size of dataset after duplicates removed:", dim(phoible)[1])

# --- Log transform variables ---
phoible$Phoneme_Count <- log(phoible$Sounds)
```

### Preview data

```{r}
head(phoible)
summary(phoible)
```

## Merging Datasets

```{r}
data <- merge(
  nee22[, 
                   c("ISO", "L1_pop", "region", "bordering_language_richness", 
                     "roughness", "altitude_range", "documentation")],
  phoible[, c("ISO639P3code", "Glottocode", "Latitude", "Longitude", "Phoneme_Count")], by.x = "ISO", by.y = "ISO639P3code", all.x = TRUE)

data <- merge(data, nee24[, c("ISO693.3", "Island.Endemic", "Distance.to.Mainland", "Distance.to.Continent", "Range.Size..km2.")], 
              by.x = "ISO", by.y = "ISO693.3", all.x = TRUE)
```

### Data Cleanup

```{r}
# --- Remove all NA entries ---
paste("Size of initial dataset:", dim(data)[1])
old_data <- data
data <- na.omit(data) 
paste("Size of dataset with NA removed:", dim(data)[1])

# --- Remove iso duplicate entries
print("Duplicate iso entries:")
iso_duplicates <- duplicated(data$ISO)
print(data$ISO[iso_duplicates])
data <- data[!unlist(iso_duplicates),]
paste("Size of dataset after duplicates removed:", dim(data[1]))
```

## Adjusting data with matrices

```{r}
common_ids <- Reduce(intersect, list(
  data$ISO,
  rownames(phylomatrix),
  rownames(spmatrix)
))
data <- data[data$ISO %in% common_ids, ]
phylomatrix <- phylomatrix[common_ids, common_ids]
spmatrix <- spmatrix[common_ids, common_ids]
```

### Preview Data

```{r}
paste("Size of dataset adjusted:", dim(data)[1])
paste("#Island_Endemic: ", sum(data$Island.Endemic))
paste("#Island_Endemic to Total Ratio: ", sum(data$Island.Endemic) / dim(data)[1])
```

### Variable Correlations

```{r}
pairs(data[,
            c("Phoneme_Count", "Range.Size..km2.",
              "L1_pop", "Distance.to.Mainland", "Distance.to.Continent",
              "altitude_range", "bordering_language_richness"
              )],
      panel = function(x, y, ...) {
        points(x, y, cex = 0.1, ...) # cex = 0.5 makes points half the default size
      })

cor(data[,
            c("Phoneme_Count", "Range.Size..km2.",
              "L1_pop", "Distance.to.Mainland", "Distance.to.Continent",
              "altitude_range", "bordering_language_richness"
              )])

cor.test(data$Phoneme_Count,data$Island.Endemic)
```

### Preview Matrices

```{r}
print(dim(phylomatrix))
print(dim(spmatrix))
print(phylomatrix[1:10,1:10])
print(spmatrix[1:10,1:10])
```

# Analysis - Generalised Least Squares (GLS) (2)

## GLS Setup (2)

```{r}
best_p <- function (p,formula,data,spmatrix,phylomatrix) {
    spmatrix <- spmatrix/max(spmatrix)
    spmatrix <- exp(-(spmatrix/p[2])^2)
    mat <- as.matrix((p[3]*(1-p[1])*spmatrix+(1-p[1])*(1-p[3])*phylomatrix+p[1]*diag(dim(phylomatrix)[1])))
    res <- try(gls(model=formula,data=data,correlation=corSymm(mat[lower.tri(mat)],fixed=T),method="ML"),silent=T)
    if (inherits(res,"try-error")) {
      out <- -10000
    } else {
      out <- res$logLik
      # if (res$logLik>0) {out <- -10000} # TESTING WITH POSITIVE LOG-LIKE
    }
    -out
}

ml_fit <- function (p,formula,data,spmatrix,phylomatrix) {
    spmatrix <- spmatrix/max(spmatrix)
    spmatrix <- exp(-(spmatrix/p[2])^2)
    mat <- as.matrix((p[3]*(1-p[1])*spmatrix+(1-p[1])*(1-p[3])*phylomatrix+p[1]*diag(dim(phylomatrix)[1])))
    res <- try(gls(model=formula,data=data,correlation=corSymm(mat[lower.tri(mat)],fixed=T),method="ML"),silent=T)
    res
}
```

### Testing GLS runtime

```{r, include=FALSE}
# Start profiling, save to file "out.prof"
Rprof("best_p.prof")

init_p_test = c(0.5,0.5,0.5)

# Run some slow code
best_p(p = init_p_test,
                 formula = Phoneme.Inventory.Size ~ 1,
                 data = nee24,
                 spmatrix = spmatrix,
                 phylomatrix = phylomatrix)

# Stop profiling
Rprof(NULL)

# Summarize results
summaryRprof("best_p.prof")
```

### Debugging GLS try-error

```{r, include=FALSE}

best_p_test <- function (p,formula,data,spmatrix,phylomatrix) {
    spmatrix <- spmatrix/max(spmatrix)
    spmatrix <- exp(-(spmatrix/p[2])^2)
    mat <- as.matrix((p[3]*(1-p[1])*spmatrix+(1-p[1])*(1-p[3])*phylomatrix+p[1]*diag(dim(phylomatrix)[1])))
    res <- tryCatch(
      gls(model = Phoneme.Inventory.Size ~ 1, data = nee24,
          correlation = corSymm(mat[lower.tri(mat)], fixed = TRUE),
          method = "ML"),
      error = function(e) e
      )
    if (inherits(res, "error")) {
      cat("gls error message:\n", res$message, "\n")
      } else {
        res$logLik
      }
}

out_test <- best_p_test(p=init_p_test,formula = Phoneme.Inventory.Size ~ 1,data = nee24,spmatrix = spmatrix,phylomatrix = phylomatrix)

print(out_test)
```

## Model predictor combinations

```{r}
predictors <- c("Island.Endemic", "Range.Size..km2.",
              "L1_pop", "Distance.to.Mainland", "Distance.to.Continent")
response <- "Phoneme_Count"
n_pred <- length(predictors)
pred_combs <- sapply(1:n_pred, function(x) combn(predictors, x))

models <- c(Phoneme.Inventory.Size~1)
for (i in 1:n_pred) {
    #print(pred_combs[[i]])
    for (j in 1:dim(pred_combs[[i]])[2]) {
        model <- (paste(response, paste(pred_combs[[i]][,j], collapse="+"), sep="~"))
        model <- as.formula(model)
        models <- c(models,model)
    }
}
unlist(models)
```

## Fitting Models

### Searching for viable initial p value

```{r}
# Get best p for all models with intercept only

# Define the parameter grid
param_values <- c(0.25, 0.5, 0.75)

# Generate all combinations
init_ps <- expand.grid(p1 = param_values, 
                       p2 = param_values, 
                       p3 = param_values)

# Convert each row to a numeric vector
init_ps_list <- split(init_ps, seq(nrow(init_ps)))
init_ps_list <- lapply(init_ps_list, function(df) as.numeric(df))

# Run mclapply over all combinations
best_p_outs <- mclapply(
  X = init_ps_list,
  FUN = function(p) best_p(p=p,
                          formula = Phoneme_Count ~ 1,
                          data = data,
                          spmatrix = spmatrix,
                          phylomatrix = phylomatrix),
  mc.cores = 7
)

print(best_p_outs)
```

### Best p values for all models

```{r}
# Get best p for all models with intercept only

# # Define the parameter grid
# param_values <- c(0.05, 0.25, 0.5, 0.75, 0.95)
# 
# # Generate all combinations
# init_ps <- expand.grid(p1 = param_values,
#                        p2 = param_values,
#                        p3 = param_values)
# 
# # Convert each row to a numeric vector
# init_ps_list <- split(init_ps, seq(nrow(init_ps)))
# init_ps_list <- lapply(init_ps_list, function(df) as.numeric(df))
# 
# # Run mclapply over all combinations
# p_fits <- mclapply(
#   X = init_ps_list,
#   FUN = function(p) sbplx(p,
#                           best_p,
#                           formula = Phoneme.Inventory.Size ~ 1,
#                           data = nee24,
#                           spmatrix = spmatrix,
#                           phylomatrix = phylomatrix,
#                           lower = c(0,0,0),
#                           upper = c(1,1,1),
#                           nl.info = TRUE),
#   mc.cores = 7
# )

p.res <- sbplx(c(0.5, 0.5, 0.5),
               best_p,
               formula=Phoneme_Count~1,
               data=data,
               spmatrix=spmatrix,phylomatrix=phylomatrix,
               lower=c(0,0,0),upper=c(1,1,1),
               nl.info = TRUE)
```

### Maximum likelihood fits for all models

```{r}

ml_fits <- mclapply(
  X = models,
  FUN = function(f) ml_fit(p=p.res$par,formula=f,data=data,spmatrix=spmatrix,phylomatrix=phylomatrix),
  mc.cores = 6
)
```

## Model Comparison

```{r}

```

# References

1.  Bromham, L., Yaxley, K.J. & Cardillo, M. Islands are engines of language diversity. *Nat Ecol Evol* **8**, 1991–2002 (2024). <https://doi.org/10.1038/s41559-024-02488-4>
2.  Hua, X., Greenhill, S.J., Cardillo, M. *et al.* The ecological drivers of variation in global language diversity. *Nat Commun* **10**, 2047 (2019). <https://doi.org/10.1038/s41467-019-09842-2>
3.  Bromham, L., Dinnage, R., Skirgård, H. *et al.* Global predictors of language endangerment and the future of linguistic diversity.
    *Nat Ecol Evol* **6**, 163–173 (2022). <https://doi.org/10.1038/s41559-021-01604-y>
4.  Cormac Anderson, Tiago Tresoldi, Simon J Greenhill, Robert Forkel, Russell Gray, Johann-Mattis List, Variation in phoneme inventories: quantifying the problem and improving comparability, *Journal of Language Evolution*, Volume 8, Issue 2, July 2023, Pages 149–168, <https://doi.org/10.1093/jole/lzad011>
