---
title: "NEE24_Analysis"
output:
  pdf_document: default
  html_document: default
date: "2025-09-19"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyr)
library(nloptr)
library(nlme)
library(parallel)
```

# Data

## Loading Phylogenetic and Spatial Distance Matrices

```{r}
phylomatrix <- read.csv("data/distance_matrices/phylogenetic_covariance_matrix.csv",
                        check.names = FALSE)
rownames(phylomatrix) <- phylomatrix[,1]
phylomatrix <- phylomatrix[,-1]
phylomatrix <- as.matrix(phylomatrix)
spmatrix <- read.csv("data/distance_matrices/spatial_distance_matrix.csv",
                     check.names = FALSE)
rownames(spmatrix) <- spmatrix[,1]
spmatrix <- spmatrix[,-1]
spmatrix <- as.matrix(spmatrix)
```

### Preview Matrices

```{r}
print(phylomatrix[1:5,1:5])
print(spmatrix[1:5,1:5])
```

## Loading NEE24 (1)

```{r}
nee24 <- read.csv("data/Islands are engines of language diversity data.csv",
                  stringsAsFactors = FALSE)
```

### NEE24 Cleanup

```{r}
# --- Remove all NA entries ---
paste("Size of initial dataset:", dim(nee24)[1])
paste("#Island_Endemic: ", sum(nee24$Island.Endemic))
paste("#Island_Endemic to Total Ratio: ", sum(nee24$Island.Endemic) / dim(nee24)[1])
old_nee24 <- nee24
nee24 <- na.omit(nee24) 
paste("Number of incomplete (NA) language entries removed:", 
      dim(old_nee24)[1] - dim(nee24)[1])
paste("Size of dataset with NA remove:", dim(nee24)[1])
paste("#Island_Endemic: ", sum(nee24$Island.Endemic))
paste("#Island_Endemic to Total Ratio: ", sum(nee24$Island.Endemic) / dim(nee24)[1])

# --- Convert variable types ---
nee24$Island.Endemic <- as.numeric(nee24$Island.Endemic)

# --- Log transform variables ---
nee24$Phoneme.Inventory.Size <- log(nee24$Phoneme.Inventory.Size) 
nee24$Range.Size..km2. <- log(nee24$Range.Size..km2.)
nee24$L1.Population <- log(nee24$L1.Population + 0.5)
nee24$Distance.to.Mainland <- log(nee24$Distance.to.Mainland + 0.5)
nee24$Distance.to.Continent <- log(nee24$Distance.to.Continent + 0.5)

```

### Preview data

```{r}
head(nee24)
summary(nee24)
```

## Adjusting NEE24 data with matrices

```{r}
common_ids <- Reduce(intersect, list(
  nee24$ISO,
  rownames(phylomatrix),
  rownames(spmatrix)
))
nee24 <- nee24[nee24$ISO %in% common_ids, ]
phylomatrix <- phylomatrix[common_ids, common_ids]
spmatrix <- spmatrix[common_ids, common_ids]
```

### Preview Data

```{r}
paste("Size of dataset adjusted:", dim(nee24)[1])
paste("#Island_Endemic: ", sum(nee24$Island.Endemic))
paste("#Island_Endemic to Total Ratio: ", sum(nee24$Island.Endemic) / dim(nee24)[1])
```

### Variable Correlations

```{r}
pairs(nee24[,
            c("Phoneme.Inventory.Size", "Range.Size..km2.",
              "L1.Population", "Distance.to.Mainland", "Distance.to.Continent")],
      panel = function(x, y, ...) {
        points(x, y, cex = 0.1, ...) # cex = 0.5 makes points half the default size
      })

cor(nee24[,
            c("Phoneme.Inventory.Size", "Range.Size..km2.",
              "L1.Population", "Distance.to.Mainland", "Distance.to.Continent")])

cor.test(nee24$Phoneme.Inventory.Size,nee24$Island.Endemic)
```

### Preview Matrices

```{r}
print(dim(phylomatrix))
print(dim(spmatrix))
print(phylomatrix[1:10,1:10])
print(spmatrix[1:10,1:10])
```

# Analysis - Generalised Least Squares (GLS) (2)

## GLS Setup (2)

```{r}
best_p <- function (p,formula,data,spmatrix,phylomatrix) {
    spmatrix <- spmatrix/max(spmatrix)
    spmatrix <- exp(-(spmatrix/p[2])^2)
    mat <- as.matrix((p[3]*(1-p[1])*spmatrix+(1-p[1])*(1-p[3])*phylomatrix+p[1]*diag(dim(phylomatrix)[1])))
    res <- try(gls(model=formula,data=data,correlation=corSymm(mat[lower.tri(mat)],fixed=T),method="ML"),silent=T)
    if (inherits(res,"try-error")) {
      out <- -10000
    } else {
      out <- res$logLik
      if (res$logLik>0) {out <- -10000}
    }
    -out
}

ml_fit <- function (p,formula,data,spmatrix,phylomatrix) {
    spmatrix <- spmatrix/max(spmatrix)
    spmatrix <- exp(-(spmatrix/p[2])^2)
    mat <- as.matrix((p[3]*(1-p[1])*spmatrix+(1-p[1])*(1-p[3])*phylomatrix+p[1]*diag(dim(phylomatrix)[1])))
    res <- try(gls(model=formula,data=data,correlation=corSymm(mat[lower.tri(mat)],fixed=T),method="ML"),silent=T)
    res
}
```

## Model predictor combinations

```{r}
predictors <- c("Island.Endemic", "Range.Size..km2.",
              "L1.Population", "Distance.to.Mainland", "Distance.to.Continent")
response <- "Phoneme.Inventory.Size"
n_pred <- length(predictors)
pred_combs <- sapply(1:n_pred, function(x) combn(predictors, x))

models <- c(Phoneme.Inventory.Size~1)
for (i in 1:n_pred) {
    #print(pred_combs[[i]])
    for (j in 1:dim(pred_combs[[i]])[2]) {
        model <- (paste(response, paste(pred_combs[[i]][,j], collapse="+"), sep="~"))
        model <- as.formula(model)
        models <- c(models,model)
    }
}
unlist(models)
```

## Fitting Models

### Best p values for all models

```{r, eval=FALSE}
# Get best p for all models with intercept only

# Define the parameter grid
param_values <- c(0.05, 0.25, 0.5, 0.75, 0.95)

# Generate all combinations
init_ps <- expand.grid(p1 = param_values, 
                       p2 = param_values, 
                       p3 = param_values)

# Convert each row to a numeric vector
init_ps_list <- split(init_ps, seq(nrow(init_ps)))
init_ps_list <- lapply(init_ps_list, function(df) as.numeric(df))

# Run mclapply over all combinations
p_fits <- mclapply(
  X = init_ps_list,
  FUN = function(p) sbplx(p,
                          best_p,
                          formula = Phoneme.Inventory.Size ~ 1,
                          data = nee24,
                          spmatrix = spmatrix,
                          phylomatrix = phylomatrix,
                          lower = c(0,0,0),
                          upper = c(1,1,1),
                          nl.info = TRUE),
  mc.cores = 8
)


p_fits <- mclapply(
  X = init_ps,
  FUN = function(p) sbplx(p,
               best_p,
               formula=Phoneme.Inventory.Size~1,
               data=nee24,
               spmatrix=spmatrix,phylomatrix=phylomatrix,
               lower=c(0,0,0),upper=c(1,1,1),
               nl.info = TRUE),
  mc.cores = 8
)

# p.res <- sbplx(c(0.5, 0.5, 0.5),
#                best_p,
#                formula=Phoneme.Inventory.Size~1,
#                data=nee24,
#                spmatrix=spmatrix,phylomatrix=phylomatrix,
#                lower=c(0,0,0),upper=c(1,1,1),
#                nl.info = TRUE)
```

### Maximum likelihood fits for all models

```{r, eval=FALSE}

ml_fits <- mclapply(
  X = models,
  FUN = function(f) ml_fit(p=p.res$par,formula=f,data=nee24,spmatrix=spmatrix,phylomatrix=phylomatrix),
  mc.cores = 6
)
```

## Model Comparison

```{r}

```

# References

1.  Bromham, L., Yaxley, K.J. & Cardillo, M. Islands are engines of language diversity. *Nat Ecol Evol* **8**, 1991–2002 (2024). <https://doi.org/10.1038/s41559-024-02488-4>
2.  Hua, X., Greenhill, S.J., Cardillo, M. *et al.* The ecological drivers of variation in global language diversity. *Nat Commun* **10**, 2047 (2019). <https://doi.org/10.1038/s41467-019-09842-2>
