---
title: "NEE24_Summary_2019"
author: "Samuel Ingloff-Richards"
date: "2025-09-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Preview data and matrices

```{r}
nee24 <- read.csv("output/nee24/nee24_adjusted.csv")
load("output/nee24/phylomatrix.RData")
load("output/nee24/spmatrix.RData")
ml_data <- read.csv("output/nee24/ml_data.csv")
```

## nee24 (4) Dataset

```{r}
dim(nee24)
print(nee24[1:6,],width = 150)
```

## Spatial Distance (3) Matrix

```{r}
paste(dim(spmatrix))
paste("Is symmetric? ", isSymmetric(spmatrix))
print(spmatrix[1:6,1:6])
```

## Phylogenetic (3) Matrix

```{r}
print(dim(phylomatrix))
paste("Is symmetric? ", isSymmetric(phylomatrix))
print(phylomatrix[1:6,1:6])
```

# Variable Table

# Method Overview

1.  Merge data from (1), (3) and (4).

2.  Remove NA entries in merged data.

3.  Remove duplicate languages (due to Phoible having multiple inventories for same language, or many Glottocodes for ISO639-3).

4.  Adjust size of matrices with merged data to only shared languages, ending up with 762 languages.

5.  Log transform variables. Although zero values in `Distance.to.Continent` and `Distance.to.Mainland` were not log transformed, e.g:

`nee24$Distance.to.Mainland[nee24$Distance.to.Mainland != 0] <- log( nee24$Distance.to.Mainland[nee24$Distance.to.Mainland != 0])`

6.  Find optimal p vector (encoding weighting of spatial and phylogenetic matrices) using 2019 (2) method, on interceot only model: `Phoneme.Inventory.Size ~ 1`
7.  Create all model combinations for given predictors, here 128 models (including intercept only model).
8.  Using 2019 (2) method and optimal p vector from step 6, find generalised least squares (GLS) maximum likelihood solution for each model.
9.  Sort models based on BIC values.

# Results using 2019 (2) Method

## First 10 models by increasing BIC

```{r}
print(ml_data[1:10,],width = 150)
```

*32 models ranked by BIC. logLik = Log likelihood, MSE = Mean Square Error (of model residuals). NA indicates that predictor was not used in the model, all NA model is the intercept model: Phoneme.Inventory.Size \~ 1. Predictor columns show their respective coefficient values, if "\*\*\*" is appended to a coefficient it indicates that the predictor in this model has a p-value below 0.05.*

## Summary

# References

1.  Bromham, L., Yaxley, K.J. & Cardillo, M. Islands are engines of language diversity. *Nat Ecol Evol* **8**, 1991–2002 (2024). <https://doi.org/10.1038/s41559-024-02488-4>
2.  Hua, X., Greenhill, S.J., Cardillo, M. *et al.* The ecological drivers of variation in global language diversity. *Nat Commun* **10**, 2047 (2019). <https://doi.org/10.1038/s41467-019-09842-2>
3.  Bromham, L., Dinnage, R., Skirgård, H. *et al.* Global predictors of language endangerment and the future of linguistic diversity. *Nat Ecol Evol* **6**, 163–173 (2022). <https://doi.org/10.1038/s41559-021-01604-y>
4.  Cormac nee24, Tiago Tresoldi, Simon J Greenhill, Robert Forkel, Russell Gray, Johann-Mattis List, Variation in phoneme inventories: quantifying the problem and improving comparability, *Journal of Language Evolution*, Volume 8, Issue 2, July 2023, Pages 149–168, <https://doi.org/10.1093/jole/lzad011>
